Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-02-06 content.tex

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}
The default option for the diversification model $p_T$ is the standard birth-death model.
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology.

@@

So far we have measured the inference error on a tree
generated according to a known (and standard) tree diversification model. 
The goal of \verb;pirouette; is to measure the expected impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree diversification model, assumed to be closely related to the Yule tree prior.
We'll use the diversity-dependent tree model \cite{etienne2011diversity},
which is a birth-death model with a speciation that is dependent on the
number of species.
We measure both the baseline 
error (that serves as a control) and the full inference error, because their 
difference shows the impact of the tree model.
We obtain the baseline error by using a twin tree (see Subsection~\ref{subsec:twinning}) for 
both the generative and best candidate models.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and the best candidate model?"

>>>>

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}
The default option for the diversification model $p_T$ is the standard birth-death model.
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. We simulate the new branching times using the TESS package (\cite{TESS, hohna2016tess}).

@@

So far we have measured the inference error on a tree
generated according to a known (and standard) tree diversification model. 
The goal of \verb;pirouette; is to measure the expected impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree diversification model, assumed to be closely related to the Yule tree prior.
We will use the diversity-dependent tree model \cite{DDD, etienne2011diversity},
which is a birth-death model with a speciation that is dependent on the
number of species.
We measure both the baseline 
error (that serves as a control) and the full inference error, because their 
difference shows the impact of the tree model.
We obtain the baseline error by using a twin tree (see Subsection~\ref{subsec:twinning}) for 
both the generative and best candidate models.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and the best candidate model?"
